Local Spectral Clustering of Density Upper Level Sets

### 2021-09-01 arxiv/examples_section.tex
Cleaned out arxiv folder.

%\newtheoremstyle{aldenthm}
%{6pt} % Space above
%{6pt} % Space below
%{\itshape} % Body font
%{} % Indent amount
%{\bfseries} % Theorem head font
%{.} % Punctuation after theorem head
%{.5em} % Space after theorem head
%{} % Theorem head spec (can be left empty, meaning `normal')

%\theoremstyle{aldenthm}
%\newtheorem{theorem}{Theorem}
%\newtheorem{definition}{Definition}
%\newtheorem{lemma}{Lemma}
%\newtheorem{proposition}{Proposition}
%\newtheorem{corollary}{Corollary}

%\newtheoremstyle{aldenrmrk}
%{6pt} % Space above
%{6pt} % Space below
%{} % Body font
%{} % Indent amount
%{\itshape} % Theorem head font
%{.} % Punctuation after theorem head
%{.5em} % Space after theorem head
%{} % Theorem head spec (can be left empty, meaning `normal')
%
%\theoremstyle{aldenrmrk}
%\newtheorem{remark}{Remark}
%%%%%% End Alden

%\title{Local Spectral Clustering of Density Upper Level Sets}

% The \author macro works with any number of authors. There are two commands
% used to separate the names and addresses of multiple authors: \And and \AND.
%
% Using \And between authors leaves it to LaTeX to determine where to break the
% lines. Using \AND forces a line break at that point. So, if LaTeX puts 3 of 4
% authors names on the first line, and the last on the second line, try using
% \AND instead of \And before the third author name.
%
%\author{
%Alden Green \And
%Sivaraman Balakrishnan \And
%Ryan J. Tibshirani}
%
%\begin{document}
%\maketitle
%
%\vspace{0.1in}
%\begin{abstract}

@@

\section{Examples}
\label{sec:examples}

In this section, we demonstrate the implications of our theory, and compare the performance of PPR density clustering to some alternatives, by walking through a series of examples. In all of these examples, the generative model under consideration will be a mixture of uniform distributions, plus some noise distribution. To formally define this model, we specify a support $\mathcal{X}$ and disjoint subsets $\mathcal{C}_1,\ldots,\mathcal{C}_M$ of $\mathcal{X}$, where $M \geq 2$ is the number of mixture components. We sample the background noise uniformly over $\mathcal{C}_0 = \mathcal{X} \setminus \bigcup_{m = 1}^{M} \mathcal{C}_m$, and let $\epsilon \in [0,1]$ be the fraction of background noise. The mixture distribution $\Pbb$ is therefore
\begin{equation*}
\Pbb = \frac{1 - \epsilon}{M} \sum_{m = 1}^{M} \Psi_m + \epsilon \Psi_0,
\end{equation*}
where $\Psi_m$ is the uniform distribution over $\Cset_m$.

Note that when $\epsilon$ is sufficiently small, for an appropriate choice of threshold $\lambda$ the density clustering will be $\Cbb_f(\lambda) = \set{\Cset_1,\ldots,\Cset_M}$. In this setup, therefore, estimating a density cluster is equivalent to recovering a mixture component.
\subsection{1d uniform example.}

We show the performance of our bounds in a mixture of uniforms in $\Reals$, and compare with the bound obtained via a global spectral clustering approach.

\subsection{General uniform example.}

We show the performance of our bounds in a mixture of uniforms in $\Reals^d$, and compare with the bound obtained via a density estimation approach.

\subsection{Lower bound.}

The theory we have developed so far gives tighter guarantees for plug-in density cluster estimators than for PPR. However, in both cases these are merely upper bounds on the clustering error. Now, we provide accompanying lower bounds for the PPR algorithm. These will demonstrate that there exist densities for which consistent estimation of density clusters is still possible, but for which \emph{PPR} will provably fail to recover density clusters (with high probability.) Moreover, they will show that the difficulty function $\kappa(\Cset)$ exhibits the right dependence on (some of) the geometric parameters we have introduced.

Our hard case will be a mixture of $2$-uniforms with background noise. Given $0 < \sigma < \rho$, let $\mathcal{X} = [-1.5\sigma,1.5\sigma] \times [-.5\rho,.5\rho]$, and let $\Cset_1 = [-1.5\sigma,-.5\sigma] \times [.5\rho,1.5\rho]$, $\Cset_2 = [.5\sigma,1.5\sigma] \times [-.5\rho,.5\rho]$, so that the background noise is sampled over $\Cset_0 = [-.5\sigma,.5\sigma] \times [-.5\rho,.5\rho]$. We denote the resulting model by $\Pbb(\sigma,\rho,\epsilon)$ to emphasize the dependence of the distribution on the parameters of interest.

\section{Technical results.}

\subsection{Lower bound.}

Let $\Sset = [-1.5\sigma,1.5\sigma] \times [-\rho/2,0]$. Let the \emph{cut} and \emph{volume} functionals be defined for $\Aset \subset \Rd$ as follows:
\begin{equation*}
\cut_{\Pbb,r}(\Aset) := \int_{\Aset} \int_{\mathcal{X} \setminus \Aset} \1(\norm{x - y} \leq r) \,d\Pbb(x) \,d\Pbb(y), \vol_{\Pbb,r}(\Aset) := \int_{\Aset} \int_{\mathcal{X}} \1(\norm{x - y} \leq r) \,d\Pbb(x) \,d\Pbb(y)
\end{equation*}
and the \emph{normalized cut},
\begin{equation*}
\Phi_{\Pbb,r}(\Aset) := \frac{\cut_{\Pbb,r}(\Aset)}{\min\{\vol_{\Pbb,r}(\Aset),\vol_{\Pbb,r}(\mathcal{X} \setminus \Aset)\}}.
\end{equation*}

\textbf{1.} We first upper bound the normalized cut $\Phi_{\Pbb,r}(\Sset)$ and lower bound the normalized cut $\Phi_{\Pbb,r}(\Cset_1)$, using the geometric parameters $\sigma$ and $\rho$, the \textcolor{red}{drop parameter} $\epsilon$, and the radius $r$. Our bounds will be crude, but will display the right dependence on these parameters, and so will be sufficient for our purposes.

We begin with the upper bound on $\Phi_{\Pbb,r}(\Sset)$. Note that only points $x = (x_1,x_2)$ such that $-r < x_2 < 0$ contribute to the cut of $\Sset$. Letting $f$ be the density function of $\Pbb$, we have
\begin{align*}
\cut_{\Pbb,r}(\Sset) & \leq \Pbb(\set{x \in \mathcal{X}: -r < x_2 < 0}) \cdot \max_{x \in \mathcal{X}} \Pbb(B(x,r)) \\
& \leq \frac{3 r}{\rho} \cdot \frac{\nu_d r^d}{\sigma \rho}.
\end{align*}
By symmetry, $\vol_{\Pbb,r}(\Sset) = \vol_{\Pbb,r}(\mathcal{X} \setminus \Sset)$, and therefore it is sufficient to lower bound $\vol_{\Pbb,r}(\Sset)$. We have
\begin{align*}
\vol_{\Pbb,r}(\Sset) & \geq \Pbb(\set{x \in \Cset_1 \cap \Sset: \dist(x, \partial \Cset_1) > r}) \cdot \frac{\nu_d r^d}{\sigma \rho} \\
& = \frac{(\sigma - 2r)(\rho - r)}{2 \sigma \rho} \cdot \frac{\nu_d r^d}{\sigma \rho}  \\
& \geq \frac{3\nu_d r^d}{16\sigma \rho},
\end{align*}
where the last inequality follows since $r \leq \sigma/4$ and $\sigma < \rho$. Therefore,
\begin{equation*}
\Phi_{\Pbb,r}(\Sset) \leq 16 \frac{r}{\rho}.
\end{equation*}
Similar manipulations yield a lower bound on $\Phi_{\Pbb,r}(\Cset_1)$. We use our estimates of the volume of spherical caps from \textcolor{red}{A.6} to obtain a lower bound on $\cut_{\Pbb,r}(\Cset_1)$, 
\begin{align*}
\cut_{\Pbb,r}(\Cset_1) & \geq \Pbb(\set{x \in \Cset_1: \dist(x,\Cset_0) < r/32}) \cdot \frac{\nu(\textrm{cap}_{r}(\frac{31r}{32})) \epsilon}{2\rho\sigma} \\
& \geq \frac{r (1 - \epsilon)}{32 \sigma} \cdot \frac{3 \nu_d r^d \epsilon}{16\rho\sigma}.
\end{align*}
and we can immediately upper bound the volume of $\Cset_1$,
\begin{equation*}
\vol_{\Pbb,r}(\Cset_1) \leq (1 - \epsilon) \cdot  \frac{(1 - \epsilon)}{\rho \sigma} \nu_d r^d.
\end{equation*}
Observe that $\vol_{\Pbb,r}(\mathcal{X} \setminus \Cset_1) \geq \vol_{\Pbb,r}(\Cset_2) = \vol_{\Pbb,r}(\Cset_1)$. Therefore,
\begin{equation}
\label{eqn:lb_proof_2}
\Phi_{\Pbb,r}(\Cset_1) \geq  \frac{\epsilon r}{256 \sigma}.
\end{equation}

\textbf{2.} The upper bound \eqref{eqn:lb_proof_2} can be extended to hold for all sets $\Aset$ with small symmetric set difference $\Aset \triangle \Cset_1$. Formally,
\begin{equation}
\label{eqn:continuity_normalized_cut}
\Phi_{\Pbb,r}(\Aset) \geq \frac{1}{2\pi} \cdot \frac{\epsilon^2r}{\sigma} \bigl(1 - 4 \cdot \Pbb(\Aset \triangle \Cset_1 )\bigr)
\end{equation}
The proof of \eqref{eqn:continuity_normalized_cut} is somewhat involved, and we defer it to the following section.

\textbf{3.} We now convert from population to sample normalized cut. \textit{Claim 1:} There exist constants $c_1,c_2$ which may depend on $(\epsilon, \rho, \sigma)$ such that the following statements holds with respect to $c_1,c_2$: For all $n$ such that
\begin{equation*}
c_1 \cdot \left(\frac{\log n}{n}\right)^{1/d} \leq r,
\end{equation*}
the sample normalized cut of $\Aset$ is lower bounded
\begin{equation*}
\Phi_{n,r}(\Aset) \geq \frac{\Phi_{\Pbb,r}(\Aset)}{4} \quad \textrm{for all $\Aset \subset \Rd$,}
\end{equation*}  
and the sample normalized cut of $\Sset$ is upper bounded,
\begin{equation*}
\Phi_{n,r}(\Sset) \leq \frac{\Phi_{\Pbb,r}(\Sset)}{2}
\end{equation*}
with probability at least $1 - \frac{c_2}{n}$.

\textbf{4.}

We prove this claim in a following section.

\subsubsection{Proof of \eqref{eqn:continuity_normalized_cut}}
To prove \eqref{eqn:continuity_normalized_cut}, we will need to upper bound the cut of $\Aset$, and lower bound the volume. We begin by upper bounding the cut, which will be the more difficult of the two. To obtain the desired result, we will need to discretize $\Aset$. For simplicity, without loss of generality assume $\rho$ and $\sigma$ are constant multiples of $r$. For $i = 0,\ldots,\frac{6\sigma}{r} - 1$ and $j = 0,\ldots, \frac{2\rho}{r} - 1$, let
\begin{equation*}
B_{ij} = \left[\frac{-3\sigma + ir}{2}, \frac{-3\sigma + (i + 1)r}{2}\right] \times \left[\frac{-\rho + jr}{2}, \frac{-\rho + (j + 1)r}{2}\right] 
\end{equation*}
and let $\mathfrak{B} = \set{B_{ij}}$ be the collection of all such boxes. Our discrete version of $\mathcal{A}$ will be a subsets of $\mathfrak{B}$,
\begin{equation*}
\overline{\mathcal{A}} := \set{B_{ij} \in \mathfrak{B}: \Pbb(B_{ij} \cap \mathcal{A}) \geq \frac{\Pbb(B_{ij})}{2}}.
\end{equation*}
(Note that $\mathcal{C}_1 = \bigcup \overline{\Cset}_1$ where $\overline{\Cset}_1 = \set{B_{ij}: i \in 0,\ldots,\frac{2\sigma}{r} - 1 }$, and is therefore effectively already discretized.)
The discretization of $\Aset$ is useful for several reasons. First, we use it to lower bound the cut of the continuous set $\Aset$. Define the boundary of the discretized set $\overline{\mathcal{A}}$ to be 
\begin{equation*}
\overline{\partial \Aset} = \set{B_{ij} \in \overline{\mathcal{A}}: \exists B_{lk} \in \mathfrak{B} ~\textrm{such that}~\abs{l - i} + \abs{k - j} = 1, B_{lk} \not\in \overline{\mathcal{A}}}.
\end{equation*}
All boxes in the boundary $\overline{\partial \Aset}$ contribute substantially to the cut of $\Aset$,
\begin{equation}
\label{eqn:lb_proof}
\cut_{\Pbb,r}(\mathcal{\Aset}) \geq \sum_{B_{ij} \in \overline{\partial\mathcal{A}}} \frac{\Pbb(B_{ij}) \cdot \min_{k,l} \Pbb(B_{kl})}{4}
\end{equation}
Moreover, the symmetric set difference $\Aset \triangle \Cset_1$ can be related to the symmetric set difference $\overline{\Aset} \triangle \overline{\Cset}_1$,
\begin{equation}
\label{eqn:lb_proof_1}
\Pbb(\Aset \triangle \Cset_1) \geq \sum_{B \in \overline{\Aset} \triangle \overline{\Cset}_1} \frac{\Pbb(B_{ij})}{2}.
\end{equation}
Next, divide $\mathcal{X}$ into rectangles horizontally, $\mathfrak{R}_j = \set{B_{ij}: i = 0,\ldots,\frac{6\sigma}{r} - 1}$. We divide our analysis into cases. For each $j$, one of
\begin{enumerate}[(i)]
	\item $\overline{\Aset} \cap \mathfrak{R}_j = \emptyset$, and therefore $(\overline{\Aset} \triangle \overline{\Cset}_1) \cap \mathfrak{R}_j = \overline{\Cset}_1 \cap \mathfrak{R}_j$, or
	\item $\overline{\Aset} \cap \mathfrak{R}_j = \mathfrak{R}_j$, and therefore $(\overline{\Aset} \triangle \overline{\Cset}_1) \cap \mathfrak{R}_j \supset \overline{\Cset}_2 \cap \mathfrak{R}_j$, or
	\item $\overline{\partial\mathcal{A}} \cap \mathfrak{R}_j \neq \emptyset$.
\end{enumerate}
occurs. Moreover, (a) for every row $\mathfrak{R}_j$, $\abs{\overline{\Cset}_1 \cap \mathfrak{R}_j} =  \abs{\overline{\Cset}_2 \cap \mathfrak{R}_j} = \frac{2\sigma}{r}$, and (b) for every $B_{ij} \in \overline{\Cset}_1 \cup \overline{\Cset}_2$, the probability
\begin{equation*}
\Pbb(B_{ij}) \geq  \frac{r^2(1 - \epsilon)}{4\rho \sigma}.
\end{equation*}
Therefore, letting $J_1 = \set{j: \textrm{(i) occurs}}$, and similarly $J_2 = \set{j: \textrm{(ii) occurs}}$ and  $J_3 = \set{j: \textrm{(iii) occurs}}$, we have from \eqref{eqn:lb_proof_1} that
\begin{align*}
\Pbb(\Aset \triangle \Cset_1) & \geq \sum_{j = 0}^{\frac{2\rho}{r} - 1} \sum_{B \in (\overline{\Aset} \triangle \overline{\Cset}_1) \cap \mathfrak{R}_j} \frac{\Pbb(B_{ij})}{2} \\
& \geq \left(\abs{J_1} + \abs{J_2}\right) \cdot \frac{\sigma}{r} \cdot \frac{r^2(1 - \epsilon)}{4\rho \sigma} \\
& = \left(\frac{2\rho}{r} - \abs{J_3}\right) \frac{r(1 - \epsilon)}{4\rho}
\end{align*}
Rearranging, we obtain
\begin{equation*}
\abs{J_3} \geq \frac{2\rho}{r}\left(1 - \frac{2\Pbb(\Aset \triangle \Cset_1)}{1 - \epsilon}\right) \geq \frac{2\rho}{r}\left(1 - 4\Pbb(\Aset \triangle \Cset_1)\right)
\end{equation*}
since \textcolor{red}{$\epsilon < 1/2$.} From \eqref{eqn:lb_proof} and the prior inequality, we have
\begin{align}
\cut_{\Pbb,r}(\Aset) & \geq \sum_{j = 0}^{\frac{2\rho}{r} - 1} \sum_{B_{ij}} \frac{\Pbb(B_{ij}) \cdot \min_{k,l} \Pbb(B_{kl})}{4} \nonumber \\
& \geq \abs{J_3} \frac{r^4\epsilon^2}{16\rho^2\sigma^2} \nonumber \\
& \geq \frac{\epsilon r}{2\sigma} \cdot \frac{r^2 \epsilon}{\rho\sigma} \cdot \bigl(1 - 4\Pbb(\Aset \triangle \Cset_1)\bigr) \label{eqn:lb_proof_3}
\end{align}

We provide the following very crude upper bound on $\vol_{\Pbb,r}(\Aset)$,
\begin{equation*}
\vol_{\Pbb,r}(\Aset) \leq \vol_{\Pbb,r}(\mathcal{X}) \leq \frac{\pi r^2}{\rho \sigma},
\end{equation*}
which along with~\eqref{eqn:lb_proof_3} results in an upper bound on the normalized cut of $\Aset$,
\begin{equation*}
\Phi_{\Pbb,r}(\Aset) \leq \frac{1}{2\pi} \cdot \frac{\epsilon^2r}{\sigma} \bigl(1 - 4 \cdot \Pbb(\Aset \triangle \Cset_1 )\bigr).
\end{equation*}

\subsubsection{Proof of Claim 1.}
\end{document}

>>>>

@@

\section{Lower bound.}
\label{sec:lower_bound.}

To show a lower bound for density clustering using PPR, we exhibit a hard case: that is, a distribution $\Pbb$ for which PPR is unlikely to recover a density cluster. Let $\mathcal{C}_{\sigma}^{(0)}$, $\mathcal{C}_{\sigma}^{(1)}$, and $\mathcal{C}_{\sigma}^{(2)}$ be rectangles in $\Reals^2$, 
\begin{equation*}
\mathcal{C}_{\sigma}^{(0)} = \biggl[-\frac{\sigma}{2}, \frac{\sigma}{2}\biggr] \times \biggl[-\frac{\rho}{2}, \frac{\rho}{2}\biggr], \quad \mathcal{C}_{\sigma}^{(1)} = \mathcal{C}_{\sigma}^{(0)} - \set{(\sigma,0)}, \quad \mathcal{C}_{\sigma}^{(2)} = \mathcal{C}_{\sigma}^{(0)} + \set{(\sigma,0)} \tag{$0 < \sigma < \rho$}
\end{equation*}
and let $\mathbb{P}$ be the mixture distribution over $\mathcal{X} = \mathcal{C}_{\sigma}^{(0)} \cup \mathcal{C}_{\sigma}^{(1)} \cup \mathcal{C}_{\sigma}^{(2)}$ given by
\begin{equation*}
\mathbb{P} = \frac{1 - \epsilon}{2} \Psi_1 + \frac{1 - \epsilon}{2} \Psi_2 + \frac{\epsilon}{2} \Psi_0,
\end{equation*}
where $\Psi_m$ is the uniform distribution over $\mathcal{C}_m$ for $m = 0,1,2$. 
The density function $f$ of $\Pbb$ is simply
\begin{equation}
\label{eqn:lb_density}
f(x) = \frac{1}{\rho\sigma}\left(\frac{1 - \epsilon}{2}\1(x \in \mathcal{C}_{\sigma}^{(1)}) + \frac{1 - \epsilon}{2}\1(x \in \mathcal{C}_{\sigma}^{(2)}) + \frac{\epsilon}{2}\1(x \in \mathcal{C}_{\sigma}^{(0)})  \right)
\end{equation}
so that for any $\epsilon < \lambda < (1 - \epsilon)/2$, $\mathbb{C}_{f}(\lambda) = \set{\mathcal{C}_{\sigma}^{(1)}, \mathcal{C}_{\sigma}^{(2)}}$.

\subsection{Lower bound for PPR.}

As the following theorem demonstrates, even when Algorithm~\textcolor{red}{1} is reasonably initialized, if the density cluster $\mathcal{C}_{\sigma}^{(1)}$ is sufficiently geometrically ill-conditioned the cluster estimator $\widehat{C}$ will fail to recover $\mathcal{C}_{\sigma}^{(1)}$. Let
\begin{equation}
\label{eqn:lower_set}
\mathcal{L} = \set{(x_1,x_2) \in \mathcal{X}: x_2 < 0}.
\end{equation}
\begin{theorem}
	\label{thm:ppr_lb}
	Suppose $r < \frac{1}{40}\rho \wedge \frac{1}{4}\sigma$,  $\alpha = 65 \Phi_{\Pbb}(\mathcal{L})$, and $(L,U) = (0,1)$ are inputs to Algorithm~\textcolor{red}{1}. Then, for any
	\begin{equation}
	\label{eqn:lb_sample_size}
	n \geq \max\set{\frac{64}{\epsilon^2 \rho \sigma \pi r^2}, \frac{8}{\epsilon}}
	\end{equation}
	the following statement holds: there exists a set $C^{g} \subset \Xbf$ with $\vol(C^g \cap \mathcal{C}_{\sigma}^{(1)}[\Xbf]) \geq \frac{1}{10}\vol_{n,r}(\mathcal{C}_{\sigma}^{(1)}[\Xbf])$ such that for any seed node $v \in C^g$, the estimator $\widehat{C}$ computed by Algorithm~\textcolor{red}{1} has symmetric set difference with $\Csig^{(1)}[\Xbf]$ of volume at least
	\begin{equation}
	\label{eqn:ppr_lb}
	\frac{\sigma \rho}{r^2}\frac{\vol_{n,r}(\widehat{C} \vartriangle \mathcal{C}_{\sigma}^{(1)}[\Xbf])}{n^2} \geq \frac{1}{4} -  c \frac{\sqrt{\frac{\sigma}{\rho}}}{\epsilon^2} \sqrt{ \log\left(\frac{\rho \sigma}{\epsilon^2 r^2}\right)\frac{\sigma}{r}}
	\end{equation}
	with probability at least $1 - c_1 n \exp\set{-c_2n}$, where $c$ is a universal constant and $c_1,c_2$ are constants which do not depend on $n$.
	Consequently, if
	\begin{equation*} 
	\epsilon^2 > \frac{c}{8} \sqrt{\frac{\sigma}{\rho}} \cdot \sqrt{ \log\left(\frac{\rho \sigma}{\epsilon^2 r^2}\right)\frac{\sigma}{r}}
	\end{equation*}
	then with high probability $\frac{\sigma \rho}{r^2}\frac{\vol_{n,r}(\widehat{C} \vartriangle \mathcal{C}_{\sigma}^{(1)}[\Xbf])}{n^2}$ is at least $1/8$. 
\end{theorem}

Note that the factor of $\frac{\sigma}{r^2}$ on the left hand side of \eqref{eqn:ppr_lb} is a constant multiple of $(1/\vol_{\Pbb,r}(\Csig^{(1)}))$; hence the left hand side is directly comparable to $\frac{\vol_{n,r}(\widehat{C} \vartriangle \mathcal{C}_{\sigma}^{(1)}[\Xbf])}{\vol_{n,r}(\Csig^{(1)}[\Xbf])}$, which is the quantity we upper bound in Theorem~\textcolor{red}{1}.

Theorem~\ref{thm:ppr_lb} is stated with respect to a particular hard case, where the density clusters are rectangular subsets of $\Reals^2$. Although we picked this setting to make the theorem statement simple, our results are easily generalized to $\Reals^d$ and to non-rectangular clusters. Additionally, although we state our lower bound with respect to PPR run on neighborhood graph, the conclusion is likely to hold for a much broader class of spectral clustering algorithms. In the proof of Theorem~\ref{thm:ppr_lb}, we rely heavily on the fact that when $\epsilon^2$ is sufficiently greater than $\frac{\sigma}{\rho}$, the normalized cut of $\Cset_1$ will be much larger than that of $\mathcal{L}$. In this case, not merely PPR but any algorithm which approximates the minimum normalized cut is unlikely to recover $\Cset_1$. Local spectral clustering algorithms based on truncated random walks \textcolor{red}{Spielman and Teng}, global spectral clustering algorithms \textcolor{red}{Shi and Malik}, and $p$-Laplacian based spectral embeddings \textcolor{red}{Buhler and Hein} all have provable upper bounds on the normalized cut of cluster they output, and thus it stands to reason that all the aforementioned approaches should struggle to estimate $\Cset_1[\Xbf]$.

\subsection{Comparison with upper bound.}

To better digest the implications of Theorem~\ref{thm:ppr_lb}, we apply our upper bound to the density function $f$ given in \eqref{eqn:lb_density}. Observe that $\Csig^{(1)}$ satisfies each of the assumptions~\textcolor{red}{(A1) - (A5).} In particular
\begin{enumerate}[label=(A\arabic*)]
	\item The density $f(x) = \frac{1 - \epsilon}{2 \rho \sigma}$ for all $x \in \Csig^{(1)}$.
	\item The clusters $\Csig^{(1)}$ and $\Csig^{(2)}$ are separated, $\dist(\Csig^{(1)}, \Csig^{(2)}) \geq \sigma$. 
	\item The density $f(x) = \frac{\epsilon}{\rho\sigma}$ for all $x$ such that $0 < \dist(x,\Csig^{(1)}) \leq \sigma$. Therefore for all such $x$,
	\begin{equation*}
	\inf_{x' \in \Csig^{(1)}} f(x') - f(x)  > \left\{\frac{1 - \epsilon}{2} - \epsilon \right\} \frac{1}{\rho \sigma}.
	\end{equation*}
	\item The set $\Csig^{(1)}$ is itself convex, and has diameter $\rho$.
	\item By symmetry, $\vol_{\Pbb,r}(\Csig^{(1)}) = \vol_{\Pbb,r}(\Csig^{(2)})$ and therefore $\vol_{\Pbb,r}(\Csig^{(1)}) \leq \frac{1}{2}\vol_{\Pbb,r}(\Reals^d)$. 
\end{enumerate}

%Technically, this density does not fit nicely into the setup of Theorem~\textcolor{red}{1}, because the sets $\Cset_{\sigma}^{(m)}$ ($ m = 1,2$) play the role of both the $\sigma$-expanded sets and the density clusters themselves. However, this truly is only a technical objection. One could modify the distribution $\Pbb$ by adding a small amount of mass to the $\sigma$-interior $\Cset^{(m)} = \set{x \in \Cset_{\sigma}^{(m)}: \dist(x,\partial\Cset_{\sigma}) \geq \sigma}$. Appropriately choosing a threshold $\lambda > \frac{(1 - \epsilon)}{2\rho\sigma}$, the resulting $\lambda$-density clusters would be $\Cbb_f(\lambda) = \set{\Cset^{(1)},\Cset^{(2)}}$, and their $\sigma$-expansions would resemble the rectangles $\Csig^{(1)}, \Csig^{(2)}$ except with slightly rounded corners. 
%Alternatively, one could directly modify Lemma~\textcolor{red}{6} -- an auxiliary Lemma, used in the proof of Theorem~\textcolor{red}{3}, which establishes an upper bound on the Lebesgue measure of $\Csig + B(0,r)$ -- to hold under the assumption $\Csig$ is a rectangle of width $\sigma$ rather than a $\sigma$-expansion. The proof of this Lemma is the only time we use the fact that $\Csig$ is a $\sigma$-expansion of a density cluster. \textcolor{red}{(Ryan and Siva.)}

\begin{remark}Technically, the rectangles $\Csig^{(1)}$ and $\Csig^{(2)}$ are not $\sigma$-expansions due to their sharp corners. To fix this, one can either modify the upper bound (specifically, Lemma~\textcolor{red}{6}) to hold with respect to rectangles of width $\sigma$, or change the sets $\Csig^{(1)}$ and $\Csig^{(2)}$ to be rectangles with rounded corners. As these details do not meaningfully change our conclusions, for simplicity we will keep our upper and lower bounds as is.
\end{remark}

If the user-specified parameters are initialized according to \textcolor{red}{(14)}, we may apply Theorem~\textcolor{red}{1}. This implies that there exists a set $\Csig^{(1)}[\Xbf] \subset \Csig^{(1)}$ with $\vol_{n,r}(\Csig[\Xbf]^g) \geq \frac{1}{2}\vol_{n,r}(\Csig[\Xbf])$ such that for any seed node $v \in \Csig^{(1)}[\Xbf]$, the estimated cluster $\widehat{C}$ output by Algorithm 1 satisfies the following upper bound:
\begin{equation*}
\vol_{n,r}(\widehat{C} \vartriangle \Csig^{(1)}[\Xbf]) \leq c \cdot \kappa(\Cset^{(1)}) \cdot \vol_{n,r}(\Csig^{(1)}[\Xbf])
\end{equation*}
where the condition number is given by
\begin{equation*}
\kappa(\Cset^{(1)}) = c \frac{\epsilon}{\sigma}\left(\frac{\rho^2}{r}\log^2\left(\frac{\Lambda_{\sigma}}{\lambda_{\sigma}^2r}\right) + c\right).
\end{equation*}

To facilitate comparisons between our upper and lower bounds, assume $\frac{1}{4}\sigma \leq \frac{1}{40}\rho$ and set $r = \frac{1}{4}\sigma$. 
For some universal constant $c > 0$, the following statements each hold with high probability:
\begin{itemize}
\item If the user-specified parameters satisfy \textcolor{red}{(14)} and for $b > 0$,
\begin{equation*}
\epsilon < c\frac{b\sigma^2}{\rho^2 \log^2\left(\frac{\Lambda_{\sigma}}{\lambda_{\sigma}^2\sigma}\right)}
\end{equation*}
then $\Delta(\widehat{C}, \mathcal{C}^{(1)}[\Xbf]) \leq b \cdot  \vol_{n,r}(\mathcal{C}^{(1)}[\Xbf])$.
\item If the user-specified parameters are given as in Theorem~\ref{thm:ppr_lb}, and 
\begin{equation*}
\epsilon > c\left({\frac{\sigma}{\rho}} \log\left(\frac{\rho \sigma}{\epsilon^2 r^2}\right)\right)^{1/4}
\end{equation*}
then $\Delta(\widehat{C}, \mathcal{C}^{(1)}[\Xbf]) \geq \frac{1}{8} \vol_{n,r}(\mathcal{C}^{(1)}[\Xbf])$.
\end{itemize}

Jointly, these upper and lower bounds give a relatively precise characterization of what it means for a density cluster to be well- or poorly-geometrically conditioned for recovery using PPR.

\begin{remark}
	It is worth pointing out that the above conclusions are reliant on specific (albeit reasonable) ranges and choices of input parameters, which in some instances differ between the upper and lower bounds. We suspect that our lower bound continues to hold even when choosing input parameters as dictated by our upper bound; however, proving this is currently beyond our technical grasp, and is a matter for future work. 
\end{remark}

%It is worth pointing out that both our lower and upper bounds are stated with respect to specific choices of the input parameters to Algorithm~\ref{alg:ppr}, and that these choices are in fact different in the case of the teleportation parameter $\alpha$ and sweep cut range $(L,U)$. At a high level, we do not believe this substantially weakens the takeaway messages of our work: if a density cluster $\mathcal{C}$ is geometrically well-conditioned, then a reasonable initialization of PPR will recover $\Csig[\Xbf]$ with low error, whereas if it is geometrically poorly-conditioned, then a different but also reasonable initialization of PPR will fail to recover $\Csig[\Xbf]$. Of course, it would be more satisfying if both our upper and lower bounds could be stated with respect to the same choices of these input parameters. Unfortunately, it is not at all obvious how to prove such a claim, as to the best of our knowledge all existing work on PPR assumes similar initialization conditions to the ones we assume; effectively, that the algorithm is well-initialized with respect to the set $S \subset G$ one is interested in recovering. We suspect that our lower bound continues to hold even when choosing input parameters as dictated by our upper bound; however, proving this is currently beyond our technical grasp, and is a matter for future work. Even so, we reiterate our belief that in tandem our upper and lower bounds represent a substantial step forward in characterizing the behavior of local clustering algorithms in a statistical context \textcolor{red}{(Ryan and Siva)}.

\begin{remark}
	We also note that our lower bound is of course specific to the PPR algorithm. As mentioned previously, density clustering is a well-studied problem, and in particular under our assumptions classical plug-in density cluster estimators can consistently recover the $\sigma$-expansion $\Csig$ of a density cluster $\Cset$, even when $\epsilon$ is large compared to $\frac{\sigma}{\rho}$. That PPR has difficulty recovering density clusters which can be easily estimated by standard plug-in approaches is not surprising, nor is it a knock on PPR. Rather, it simply reflects that while classical density clustering approaches are specifically designed to identify high-density regions regardless of their geometry, PPR considers geometry as well as density when deciding upon the optimal cluster.
\end{remark}

\clearpage

\section{Proof of Lower Bound.}

To prove Theorem~\ref{thm:ppr_lb}, we will proceeding according to the following steps:
\begin{enumerate}
	\item We study the spectral partitioning properties of PPR on an arbitrary graph $G$, and show that when suitably initialized inside a subset $S \subset V$, the normalized cut of the PPR sweep cut is upper bounded by (a function of) $\Phi(S)$. 
	\item We specialize to the graph $G = G_{n,r}$ and the subset $\mathcal{L}[\Xbf] \subset \Xbf$, and show that the normalized cut $\Phi_{n,r}(\mathcal{L}[\Xbf])$ is small (with high probability) when the diameter $\rho$ is large.
	\item We reason that for the input parameters given in Theorem~\ref{thm:ppr_lb}, the output of Algorithm~\textcolor{red}{1} $\widehat{C}$ must therefore also have small normalized cut.
	\item On other hand, we show that when the noise parameter $\epsilon$ is not too small, the empirical density cluster $\Csig^{(1)}[\Xbf]$ will have large normalized cut $\Phi_{n,r}(\Csig^{(1)}[\Xbf])$. In fact, we generalize this to hold for any set $A \subset \Xbf$ for which the symmetric set distance metric $\Delta(A,\Cset_1[\Xbf])$ is small.
	\item We conclude that the symmetric set distance metric $\Delta(\widehat{C},\Csig^{(1)}[\Xbf])$ must not be small.
\end{enumerate}

We devote the subsequent sections to proving each of the aforementioned steps.

\subsection{Spectral partitioning properties of PPR.}
